---
title: "nct_project_concept_mapping_and_stats"
output: html_document
---

```{r}


library(gtsummary)
library(tidyverse)
library(lubridate)
library(dplyr)
library(forcats)
library(tidyr)
library(data.table)
library(stringr)
library(janitor)
library(bit64)
library(naniar)
library(finalfit)
library(tools)
library(ggsci)


`%notin%` <- Negate(`%in%`)
data_path <- getwd()

```

Find MeSH Curie by extact match in Athena-OMOP Database
```{r}

clean_athena_concepts <- athena_concepts %>% 
    mutate(
      lower_concept_name = tolower(concept_name)
    ) 

parsed_mesh_diseases_from_acct <- comb_submitted_and_parsed_mesh_terms %>%
  select(parsed_mesh_string_name_lower) %>%
  distinct()

get_curies_from_parsed_mesh_diseases_from_acct  <- clean_athena_concepts %>%
  filter(
    vocabulary_id == "MeSH"
  ) %>%
  distinct() %>%
  mutate(
    omop_concept_code = concept_id 
  ) %>%
  select(
    omop_concept_code, 
    concept_code, 
    lower_concept_name
  ) %>%
  distinct() 
  



```

Import Athena Files 
  - Import
    - athena_concepts
    - athena_relationships
    - athena_synonym
    - athena_ancestor
    - athena_concept_class
  - Output
    - pubtater_disease_nct_export
```{r}


athena_concepts <- fread("/Users/michaelpatton/phd/thesis/thesis_data/2021_09_21_athena_vocab_donwload/CONCEPT.csv", 
                         sep = "\t", 
                         stringsAsFactors = F, 
                         quote="") %>%
  clean_names() %>%
  as_tibble()

athena_relationships <- fread("/Users/michaelpatton/phd/thesis/thesis_data/2021_09_21_athena_vocab_donwload/CONCEPT_RELATIONSHIP.csv", 
                              sep = "\t", 
                              stringsAsFactors = F, 
                              quote="") %>%
  clean_names() %>%
  as_tibble()

eq_paths <- athena_relationships %>%
  filter(str_detect(relationship_id," eq|Maps to")) %>%
  inner_join(
    athena_concepts %>% select(concept_id, 
                               concept_code,
                               concept_name, 
                               domain_id, 
                              vocabulary_id, 
                              concept_code), by = c("concept_id_1" = "concept_id")
  ) %>%
  mutate(
    concept_name_1 = concept_name, 
    concept_code_1 = concept_code,
    domain_id_1 = domain_id, 
    vocabulary_id_1 = vocabulary_id
  ) %>%
  select(
    concept_id_1,
    concept_name_1, 
    concept_code_1, 
    domain_id_1, 
    vocabulary_id_1,
    relationship_id, 
    concept_id_2
    ) %>%
  distinct() %>%
  inner_join(
    athena_concepts %>% select(concept_id, 
                               concept_code,
                               concept_name, domain_id, 
                              vocabulary_id, 
                              concept_code), by = c("concept_id_2" = "concept_id")
  ) %>%
  mutate(
    concept_name_2 = concept_name, 
    concept_code_2 = concept_code,
    domain_id_2 = domain_id, 
    vocabulary_id_2 = vocabulary_id
  ) %>%
  select(
    concept_id_1, 
    concept_name_1, 
    concept_code_1,
    domain_id_1, 
    vocabulary_id_1,
    relationship_id, 
    concept_id_2, 
    concept_name_2, 
    concept_code_2,
    domain_id_2,
    vocabulary_id_2
  ) %>%
  distinct()
  
  

athena_synonym <- fread("/Users/michaelpatton/phd/thesis/thesis_data/2021_09_21_athena_vocab_donwload/CONCEPT_SYNONYM.csv", 
                        sep = "\t", 
                        stringsAsFactors = F, 
                        quote="") %>%
  clean_names() %>%
  as_tibble()



athena_ancestor <- fread("/Users/michaelpatton/phd/thesis/thesis_data/2021_09_21_athena_vocab_donwload/CONCEPT_ANCESTOR.csv", 
                         sep = "\t", 
                         stringsAsFactors = F, 
                         quote="") %>%
  clean_names() %>%
  as_tibble()

athena_concept_class <-  fread("/Users/michaelpatton/phd/thesis/thesis_data/2021_09_21_athena_vocab_donwload/CONCEPT_CLASS.csv", 
                               sep = "\t", 
                               stringsAsFactors = F, 
                               quote="") %>%
  clean_names() %>%
  as_tibble()

athena_drug_strength <-  fread("/Users/michaelpatton/phd/thesis/thesis_data/2021_09_21_athena_vocab_donwload/DRUG_STRENGTH.csv", 
                               sep = "\t", 
                               stringsAsFactors = F, 
                               quote="") %>%
  clean_names() %>%
  as_tibble()

# pubtater results contained numerous errors and were ommitted from final results
test <- fread("/Users/michaelpatton/phd/phd_classes/info_795_natural_language_processing/nlp_final_project/662termproject/data/PubTator_cleaned_des_out.txt", header = F, sep = "\t", fill=TRUE) %>%
  as_tibble() %>%
  clean_names()

slice_test <- test %>%
  head(3)

tester <- test %>% 
  rowwise() %>%
  mutate(
    nct_id = str_match(v1, "^NCT[0-9A-Z]{0,8}"), 
    mesh_curie_id = str_match(v1, "MESH:[0-9A-Z]{0,8}"), 
    omim_curie_id = str_match(v1, "OMIM:[0-9A-Z]{0,8}")
  ) %>%
  ungroup() %>%
  distinct()


pubtater_disease_nct_export <- tester %>%
  group_by(nct_id) %>%
  filter(!is.na(mesh_curie_id) |
           !is.na(omim_curie_id)) %>%
  ungroup() %>%
  select(nct_id,
         mesh_curie_id,
         omim_curie_id) %>%
  distinct() %>%
  as_tibble() %>%
  pivot_longer(
    cols = c(mesh_curie_id, 
             omim_curie_id), 
    #names_to = "curies", 
    values_to = "pubtater_disease_curies"
  ) %>%
  select(
    nct_id, 
    pubtater_disease_curies
  ) %>%
  drop_na() %>%
  distinct() %>%
  group_by(nct_id) %>%
  mutate(
    n_distinct_pubtater_curies = 
      n_distinct(pubtater_disease_curies)
  ) %>%
  ungroup() %>%
  distinct() %>%
  rowwise() %>%
   mutate(
    pubtater_disease_curie_code = 
      str_split(pubtater_disease_curies, ":")[[1]][2]
  ) %>%
  ungroup() %>%
  distinct() %>%
  left_join(
    get_curies_from_parsed_mesh_diseases_from_acct, 
    by = c("pubtater_disease_curie_code" =  "concept_code")
  )




```

Import 20210601 Clinical Trials
  Output
    - comb_submitted_and_parsed_mesh_terms
```{r}

filenames_nlp_clinical_trials_20211108 <-
  list.files(path = "/Users/michaelpatton/phd/phd_classes/info_795_natural_language_processing/nlp_final_project/662termproject/20211108_pipe-delimited-export", full.names = T)

nlp_clinical_trials_20211108 <- 
  lapply(filenames_nlp_clinical_trials_20211108, function(x)
  fread(
    x,
    stringsAsFactors = F,
    sep = "|"
  ) %>%
  as_tibble())

all_trial_info <- nlp_clinical_trials_20211108[[42]]


trial_description <-  nlp_clinical_trials_20211108[[14]] %>%
  select(nct_id, description) %>%
  mutate(
    trial_description = description
  ) %>%
  select(
    nct_id, 
    trial_description
  ) %>%
  distinct()

condition_string_submitted_by_trial_manager <- nlp_clinical_trials_20211108[[8]] %>%
  mutate(
    submitted_mesh_string_name = name, 
    submitted_mesh_string_name_lower = tolower(name)
  ) %>%
  select(
    nct_id, 
    submitted_mesh_string_name, 
    submitted_mesh_string_name_lower
  ) %>%
  distinct() %>%
  group_by(nct_id) %>%
  mutate(n_distinct_submitted_strings =
           n_distinct(submitted_mesh_string_name_lower)) %>%
  ungroup() %>%
  distinct()

mesh_parsed_conditions <- nlp_clinical_trials_20211108[[4]] %>%
    mutate(
    parsed_mesh_string_name = mesh_term, 
    parsed_mesh_string_name_lower = tolower(mesh_term)
  ) %>%
  select(
    nct_id, 
    parsed_mesh_string_name, 
    parsed_mesh_string_name_lower
  ) %>%
  distinct() %>%
  group_by(nct_id) %>%
    mutate(
    n_distinct_acct_parsed_strings = 
      n_distinct(parsed_mesh_string_name_lower)
  ) %>%
  ungroup() %>%
  distinct()

intervention_string_submitted_by_trial_manager <- nlp_clinical_trials_20211108[[23]] %>%
  mutate(
    submitted_mesh_string_name = name, 
    submitted_mesh_string_name_lower = tolower(name)
  ) %>%
  select(
    nct_id, 
    submitted_mesh_string_name, 
    submitted_mesh_string_name_lower
  ) %>%
  distinct() %>%
  group_by(nct_id) %>%
  mutate(n_distinct_submitted_strings =
           n_distinct(submitted_mesh_string_name_lower)) %>%
  ungroup() %>%
  distinct()


mesh_parsed_interventions <- nlp_clinical_trials_20211108[[5]] %>%
    mutate(
    parsed_mesh_string_name = mesh_term, 
    parsed_mesh_string_name_lower = tolower(mesh_term)
  ) %>%
  select(
    nct_id, 
    parsed_mesh_string_name, 
    parsed_mesh_string_name_lower
  ) %>%
  distinct() %>%
  group_by(nct_id) %>%
    mutate(
    n_distinct_acct_parsed_strings = 
      n_distinct(parsed_mesh_string_name_lower)
  ) %>%
  ungroup() %>%
  distinct()

comb_submitted_and_parsed_mesh_terms <- condition_string_submitted_by_trial_manager %>%
  left_join(
    mesh_parsed_conditions, 
    by = "nct_id"
  ) %>%
  distinct() %>%
  left_join(
    get_curies_from_parsed_mesh_diseases_from_acct, 
    by = c("parsed_mesh_string_name_lower" = "lower_concept_name" )
  ) 

interventions_comb_submitted_and_parsed_mesh_terms <- intervention_string_submitted_by_trial_manager %>%
  left_join(
    mesh_parsed_interventions, 
    by = "nct_id"
  ) %>%
  distinct() %>%
  left_join(
    get_curies_from_parsed_mesh_diseases_from_acct, 
    by = c("parsed_mesh_string_name_lower" = "lower_concept_name" )
  ) 


```

NLP output from Cubner 
  Input
    - disease_terms_kwh
    - treatment_terms_kwh
```{r}

disease_terms_kwh <- fread("/Users/michaelpatton/phd/phd_classes/info_795_natural_language_processing/nlp_final_project/662termproject/data/NLP/disease_terms.csv", header = T, sep = ",", fill=TRUE) %>%
  as_tibble() %>%
  clean_names()

treatment_terms_kwh <- fread("/Users/michaelpatton/phd/phd_classes/info_795_natural_language_processing/nlp_final_project/662termproject/data/NLP/treatment_terms.csv", header = T, sep = ",", fill=TRUE) %>%
  as_tibble() %>%
  clean_names()

```

Disease Term String Matching
  - disease_terms_kwh
  Output 
    - disease_synonyms_after_match
```{r}

disease_term_to_lower <- disease_terms_kwh %>%
  select(disease) %>%
  mutate(disease = tolower(disease)) %>%
  distinct()

disease_term_to_lower_ls <- disease_term_to_lower$disease %>% unique()

comb_submitted_and_parsed_mesh_terms_ls <- comb_submitted_and_parsed_mesh_terms$submitted_mesh_string_name_lower %>% unique()

#disease_term_to_lower_ls_short <- disease_term_to_lower_ls[1:100]

exact_match_disease_term_to_athena <- clean_athena_concepts %>%
  filter(lower_concept_name %in% disease_term_to_lower_ls |
        lower_concept_name %in% comb_submitted_and_parsed_mesh_terms_ls) %>%
  select(
    concept_name,
    lower_concept_name,
    concept_id,
    domain_id,
    vocabulary_id,
    concept_class_id,
    concept_code
  ) %>%
  distinct()

## testing
# disease_NCT05110365 <- disease_terms_kwh %>%
#   filter(nct_id == "NCT05110365") 
#   inner_join(
#     exact_match_disease_term_to_athena %>% select(lower_concept_name, concept_id, vocabulary_id, concept_code),
#     by = c("disease" = "lower_concept_name")
#   ) 

submitted_terms <- comb_submitted_and_parsed_mesh_terms %>% 
  mutate(
    disease = submitted_mesh_string_name_lower
  ) %>%
  select(
    nct_id, 
    disease
  ) %>%
  distinct()

  
disease_terms_kwh_comb_submitted_terms <- bind_rows(disease_terms_kwh, submitted_terms) %>% distinct() %>%
  arrange(nct_id)
  
disease_exact_matching <- disease_terms_kwh_comb_submitted_terms %>%
  inner_join(
    exact_match_disease_term_to_athena %>% select(
      lower_concept_name,
      domain_id,
      concept_id,
      vocabulary_id,
      concept_code
    ),
    by = c("disease" = "lower_concept_name"),
    keep = T
  ) %>%
  filter(
    domain_id %notin% c(
      "Geography",
      "Visit",
      "Episode",
      "Gender",
      "Unit",
      "Race",
      "Cost",
      "Drug",
      "Provider",
      "Payer",
      "Relationship",
      "Type Concept",
      "Procedure",
      "Device"
    )
  ) %>%
  group_by(nct_id) %>%
  mutate(n_distinct_concepts_exact_matched = n_distinct(lower_concept_name)) %>%
  ungroup() %>%
  distinct()


disease_synonyms_after_match <- disease_exact_matching %>%
  inner_join(
    eq_paths, 
    by = c("concept_id" = "concept_id_1")
  ) %>%
  group_by(nct_id) %>%
  mutate(n_distinct_synonyms_from_exact_matched = n_distinct(concept_name_2)) %>%
  ungroup() %>%
  distinct()
    

disease_exact_matching_stats <- disease_synonyms_after_match %>%
  select(
    nct_id, 
    n_distinct_concepts_exact_matched, 
    n_distinct_synonyms_from_exact_matched
  ) %>%
  distinct()

# 
# disease_exact_matching_ls_cols <- disease_exact_matching %>%
#   mutate(
#     combined_code = paste(vocabulary_id,concept_code, sep= ":")
#   ) %>%
#   ungroup() %>%
#   distinct() %>%
#   select(
#     nct_id, 
#     n_distinct_concepts_exact_matched, 
#     combined_code
#   ) %>%
#   group_by(nct_id) %>%
#   mutate(
#     codes_matched = list(combined_code)
#   ) %>%
#   ungroup() %>%
#   distinct() 
# 

```

Combine Stats from with disease 
  - comb_submitted_and_parsed_mesh_terms
```{r}

conditions_with_stats_and_data <- comb_submitted_and_parsed_mesh_terms %>%
  select(
    nct_id, 
    n_distinct_submitted_strings,
    n_distinct_acct_parsed_strings
  ) %>%
  distinct() %>%
  inner_join(
    disease_synonyms_after_match,
    by = "nct_id", 
    keep = T
  ) %>%
  distinct() %>%
  mutate(
    n_distinct_acct_parsed_strings = 
      ifelse(
        is.na(n_distinct_acct_parsed_strings), 
        0, 
        n_distinct_acct_parsed_strings
      )
  )

conditions_with_stats_and_data$vocabulary_id %>% unique()

conditions_with_stats <- comb_submitted_and_parsed_mesh_terms %>%
  select(
    nct_id, 
    n_distinct_submitted_strings,
    n_distinct_acct_parsed_strings
  ) %>%
  distinct() %>%
  inner_join(
    disease_synonyms_after_match %>% select(nct_id, n_distinct_concepts_exact_matched, n_distinct_synonyms_from_exact_matched),
    by = "nct_id"
  ) %>%
  distinct() %>%
  mutate(
    n_distinct_acct_parsed_strings = 
      ifelse(
        is.na(n_distinct_acct_parsed_strings), 
        0, 
        n_distinct_acct_parsed_strings
      )
  )

conditions_with_stats


improvement_of_mapping_conditions_with_stats <- conditions_with_stats %>%
  group_by(nct_id) %>%
  mutate(
    diff_acct_kwp = 
    n_distinct_concepts_exact_matched - n_distinct_acct_parsed_strings, 
    diff_acct_kwp_with_synonyms = 
      (n_distinct_concepts_exact_matched + n_distinct_synonyms_from_exact_matched) - n_distinct_acct_parsed_strings
  ) %>%
  ungroup() %>%
  distinct() 

# 46049
improvement_of_mapping_conditions_with_stats_better$nct_id %>% unique() %>% length()

improvement_of_mapping_conditions_with_stats_better <- improvement_of_mapping_conditions_with_stats %>%
  filter(
    diff_acct_kwp > 0
  )

# 1.40
improvement_of_mapping_conditions_with_stats_better %>%
  summarise(
    mean(diff_acct_kwp)
  )

improvement_of_mapping_conditions_with_stats_better_synonyms <- improvement_of_mapping_conditions_with_stats %>%
  filter(
    diff_acct_kwp_with_synonyms > 0
  )

#2.39
improvement_of_mapping_conditions_with_stats_better_synonyms %>%
  summarise(
    mean(diff_acct_kwp_with_synonyms)
  )


acct_0_parse_kwp_better_conditions_with_stats<- conditions_with_stats %>%
  filter(
    n_distinct_acct_parsed_strings == 0 & 
      n_distinct_concepts_exact_matched > 0
  ) %>%
  distinct()

#15625
acct_0_parse_kwp_better_conditions_with_stats$nct_id %>% unique() %>% length()



```

Extract Example Improvements for Conditions/Diseases
```{r}

example_nash <- conditions_with_stats_and_data %>%
  filter(
    str_detect(disease, "nash|non-alcoholic")
  )

example_nash <- treatment_exact_matching %>%
  filter(
    str_detect(disease, "nash|non-alcoholic")
  )



```

Treatment Term String Matching
  - treatment_terms_kwh
```{r}

### treatment terms 
  
treatment_term_to_lower <- treatment_terms_kwh %>%
  select(treatment) %>%
  mutate(treatment = tolower(treatment)) %>%
  distinct()

treatment_term_to_lower_ls <- treatment_term_to_lower$treatment

interventions_comb_submitted_and_parsed_mesh_terms_ls <- interventions_comb_submitted_and_parsed_mesh_terms$submitted_mesh_string_name_lower %>% unique()


exact_match_treatment_term_to_athena <- clean_athena_concepts %>%
    filter(lower_concept_name %in% treatment_term_to_lower_ls |
           lower_concept_name %in% interventions_comb_submitted_and_parsed_mesh_terms_ls) %>%
  select(
    concept_name,
    lower_concept_name,
    concept_id,
    domain_id,
    vocabulary_id,
    concept_class_id,
    concept_code
  ) %>%
  distinct()

intervention_submitted_terms <- interventions_comb_submitted_and_parsed_mesh_terms %>% 
  mutate(
    treatment = submitted_mesh_string_name_lower
  ) %>%
  select(
    nct_id, 
    treatment
  ) %>%
  distinct()

  
intervention_terms_from_kaiwen_comb_submitted_terms <- bind_rows(treatment_terms_kwh, intervention_submitted_terms) %>% distinct() %>%
  arrange(nct_id)
  

treatment_exact_matching <-
  intervention_terms_from_kaiwen_comb_submitted_terms %>%
  inner_join(
    exact_match_treatment_term_to_athena %>% select(
      lower_concept_name,
      domain_id,
      concept_id,
      vocabulary_id,
      concept_code
    ),
    by = c("treatment" = "lower_concept_name"),
    keep = T
  ) %>%
  filter(
    domain_id %notin% c(
      "Geography",
      "Visit",
      "Episode",
      "Gender",
      "Unit",
      "Race",
      "Cost",
      "Provider",
      "Payer",
      "Relationship",
      "Type Concept"
    )
  ) %>%
  group_by(nct_id) %>%
  mutate(n_distinct_concepts_exact_matched = n_distinct(lower_concept_name)) %>%
  ungroup() %>%
  distinct()

treatment_exact_matching$domain_id %>% unique() 


treatment_synonyms_after_match <- treatment_exact_matching %>%
  inner_join(
    eq_paths, 
    by = c("concept_id" = "concept_id_1")
  ) %>%
  group_by(nct_id) %>%
  mutate(n_distinct_synonyms_from_exact_matched = n_distinct(concept_name_2)) %>%
  ungroup() %>%
  distinct()
    

treatment_exact_matching_stats <- treatment_synonyms_after_match %>%
  select(
    nct_id, 
    n_distinct_concepts_exact_matched, 
    n_distinct_synonyms_from_exact_matched
  ) %>%
  distinct()



```

Combine Stats from with treatment
  - interventions_comb_submitted_and_parsed_mesh_terms
```{r}

interventions_with_stats_and_data$vocabulary_id %>% unique()

interventions_with_stats_and_data <- interventions_comb_submitted_and_parsed_mesh_terms %>%
  select(
    nct_id, 
    n_distinct_submitted_strings,
    n_distinct_acct_parsed_strings
  ) %>%
  distinct() %>%
  inner_join(
    treatment_synonyms_after_match,
    by = c("nct_id" = "nct_id")
    #keep = T
  ) %>%
  distinct() %>%
  mutate(
    n_distinct_acct_parsed_strings = 
      ifelse(
        is.na(n_distinct_acct_parsed_strings), 
        0, 
        n_distinct_acct_parsed_strings
      )
  )

interventions_with_stats <- interventions_comb_submitted_and_parsed_mesh_terms %>%
  select(
    nct_id, 
    n_distinct_submitted_strings,
    n_distinct_acct_parsed_strings
  ) %>%
  distinct() %>%
  inner_join(
    treatment_synonyms_after_match %>%select(nct_id, n_distinct_concepts_exact_matched,n_distinct_synonyms_from_exact_matched),
    by = "nct_id"
  ) %>%
  distinct() %>%
  mutate(
    n_distinct_acct_parsed_strings = 
      ifelse(
        is.na(n_distinct_acct_parsed_strings), 
        0, 
        n_distinct_acct_parsed_strings
      )
  )

improvement_of_mapping_interventions_with_stats <- interventions_with_stats %>%
  group_by(nct_id) %>%
  mutate(
    diff_acct_kwp = 
    n_distinct_concepts_exact_matched -n_distinct_acct_parsed_strings, 
    diff_acct_kwp_with_synonyms = 
      (n_distinct_concepts_exact_matched + n_distinct_synonyms_from_exact_matched) - n_distinct_acct_parsed_strings
  ) %>%
  ungroup() %>%
  distinct() 

improvement_of_mapping_interventions_with_stats_better <- improvement_of_mapping_interventions_with_stats %>%
  filter(
    diff_acct_kwp > 0
  )

# 84490
improvement_of_mapping_interventions_with_stats_better$nct_id %>% unique() %>% length()

# 1.32
improvement_of_mapping_interventions_with_stats_better %>%
  summarise(
  avg = sum(diff_acct_kwp) / n_distinct(nct_id) 
  )

improvement_of_mapping_interventions_with_stats_better_synonyms <- improvement_of_mapping_interventions_with_stats %>%
  filter(
    diff_acct_kwp_with_synonyms > 0
  )

#6.43
improvement_of_mapping_interventions_with_stats_better_synonyms %>%
  summarise(
    mean(diff_acct_kwp_with_synonyms)
  )


improvement_of_mapping_interventions_with_stats_better %>%
  summarise(
    mean(diff_acct_kwp)
  )




acct_0_parse_kwp_better_interventions_with_stats<- interventions_with_stats %>%
  filter(
    n_distinct_acct_parsed_strings == 0 & 
      n_distinct_concepts_exact_matched > 0
  ) %>%
  distinct()

#59383
acct_0_parse_kwp_better_interventions_with_stats$nct_id %>% unique() %>% length()


```
